Supp mat: time series paper

### 2017-06-09 writing/manuscript_draft.Rmd
shortened abstract and set double spacing

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
fontsize: 12pt
geometry: margin=1in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \usepackage{setspace}             
    \usepackage{lineno}
    \usepackage[round]{natbib}
    \bibpunct[; ]{(}{)}{,}{a}{}{;}
    \linenumbers
---

\begin{spacing}{1.5} %use 2
	\begin{flushleft}
Running head: 
\vspace{3 mm}

@@

\end{flushleft}
\end{spacing}

@@

\clearpage

## Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. In addition, census data is often expensive, requiring a lot of people-hours or equipment. However, little work has actually addressed the length of time series required. In other words, when is a time series of census data long enough to address a question of interest? Here, I explore two approaches to address this question: one simulation-based and an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using simple simulations, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size. In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. Here I show 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length. These results point to the importance of sampling populations over long periods of time. Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance. Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

>>>>

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
fontsize: 12pt
geometry: margin=1in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \usepackage{setspace}\doublespacing           
    \usepackage{lineno}
    \usepackage[round]{natbib}
    \bibpunct[; ]{(}{)}{,}{a}{}{;}
    \linenumbers
---

Running head: 
\vspace{3 mm}

@@

@@

## Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. However, census data is often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, I explore two approaches to address this question: one simulation-based and an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. <!--Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.--> Using a simulation approach, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size. In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. I show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

### 2017-06-09 writing/manuscript_draft.tex
shortened abstract and set double spacing

\usepackage{float} \usepackage{setspace} \usepackage{lineno}
\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

@@

\begin{spacing}{1.5} %use 2
    \begin{flushleft}
Running head: 
\vspace{3 mm}

@@

Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
\vspace{3 mm}

$^2$Corresponding author: eawhite@ucdavis.edu
\vspace{3 mm}

@@

\clearpage

@@

Long-term census data is necessary to better understand ecological
processes and make management decisions. In addition, census data is
often expensive, requiring a lot of people-hours or equipment. However,
little work has actually addressed the length of time series required.
In other words, when is a time series of census data long enough to
address a question of interest? Here, I explore two approaches to
address this question: one simulation-based and an empirical approach. I
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, I examine the ability to detect a trend with a set level of
statistical power, which is often neglected in ecological time series
analyses. Using simple simulations, I demonstrate how the minimum time
series length required increases with weaker trends in abundance and
with higher variability in population size. In addition, I examine 868
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. Here I show 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. Perhaps
surprisingly, the minimum time required did not correlate well with
biological explanatory variables, like body size or generation length.
These results point to the importance of sampling populations over long
periods of time. Here, I stress the need for verifying a study has
enough statistical power to accurately determine long-term changes in
population abundance. Most studies, especially those less than 10-15
years, are probably underpowered and potentially misleading.

>>>>

\usepackage{float}
\usepackage{setspace}\doublespacing \usepackage{lineno}
\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

@@

Running head: \vspace{3 mm}

@@

Addresses: \textbackslash{}
\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
\vspace{3 mm}

\(^2\)Corresponding author:
\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
\vspace{3 mm}

@@

@@

Long-term census data is necessary to better understand ecological
processes and make management decisions. However, census data is often
expensive, requiring a lot of people-hours and equipment. Little work
has addressed the question, when is a time series of census data long
enough to address a question of interest? Here, I explore two approaches
to address this question: one simulation-based and an empirical
approach. I specifically determine the minimum time series length
required to estimate significant increases or decreases in population
abundance. Using a simulation approach, I demonstrate how the minimum
time series length required increases with weaker trends in abundance
and with higher variability in population size. In addition, I examine
868 populations of vertebrate species to determine the minimum time
required to detect changes in their abundance. I show that 10-15 years
of continuous monitoring are required in order to achieve a high level
of statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. These results
point to the importance of sampling populations over long periods of
time. Most studies, especially those less than 10-15 years, are probably
underpowered and potentially misleading.
